\chapter{Greatest common divisors}
\epigraph[author={Euclid}]{The laws of nature are but the mathematical thoughts of God.}\SubIndex{Euclid}

\epigraph[author={Edna St. Vincent Millay}, source={Euclid Alone}]
{Euclid alone has looked on Beauty bare. \\
Let all who prate of Beauty hold their peace, \\
And lay them prone upon the earth and cease \\
To ponder on themselves, the while they stare \\
At nothing, intricately drawn nowhere \\
In shapes of shifting lineage; let geese \\
Gabble and hiss, but heroes seek release \\
From dusty bondage into luminous air. \\
O blinding hour, O holy, terrible day, \\
When first the shaft into his vision shone \\
Of light anatomized! Euclid alone \\
Has looked on Beauty bare. Fortunate they \\
Who, though once only and then but far away, \\
Have heard her massive sandal set on stone.}\SubIndex{St. Vincent Millay, Edna}



\section{The extended Euclidean algorithm}\define{extended Euclidean algorithm}\define{Euclidean algorithm!extended}\define{algorithm!extended Euclidean}


\begin{theorem}[B\'ezout]
For any two integers \(b, c\), not both zero, there are integers \(s,t\) so that \(sb+tc=\gcd{b,c}\).
\end{theorem}
These \emph{B\'ezout coefficients}\define{B\'ezout!coefficients} \(s, t\) play an essential role in advanced arithmetic, as we will see.

\begin{example}
To find the B\'ezout coefficients of \(12, 8\), write out a matrix
\[
\begin{pmatrix}
1 & 0 & 12 \\
0 & 1 & 8
\end{pmatrix}.
\]
Repeatedly add some integer multiple of one row to the other, to try to make the bigger number in the last column (bigger by absolute value) get smaller (by absolute value).
In this case, we can subtract row 2 from row 1, to get rid of as much of the \(12\) as possible.
All operations are carried out a whole row at a time:
\[
\begin{pmatrix}
1 & -1 & 4 \\
0 & 1 & 8
\end{pmatrix}
\]
Now the bigger number (by absolute value) in the last column is the \(8\).
We subtract as large an integer multiple of row 1 from row 2, in this case subtract \(2\pr{\text{row 1}}\) from row 2, to kill as much of the \(8\) as possible:
\[
\begin{pmatrix}
1 & -1 & 4 \\
-2 & 3 & 0
\end{pmatrix}
\]
Once we get a zero in the last column, the other entry in the last column is the  greatest common divisor, and the entries in the same row as the greatest common divisor are the B\'ezout coefficients.
In our example, the B\'ezout coefficients of \(12, 8\) are \(1,-1\) and the greatest common divisor is \(4\).
This trick to calculate B\'ezout coefficients is the \emph{extended Euclidean algorithm}.\define{extended Euclidean algorithm}\define{Euclidean algorithm!extended}
\end{example}

\begin{example}
Again, we have to be a little bit careful about minus signs.
For example, if we look at \(-8,-4\), we find B\'ezout coeffiicients by
\[
\begin{pmatrix}
1 & 0 & -8 \\
0 & 1 & -4
\end{pmatrix} \to
\begin{pmatrix}
1 & -2 & 0 \\
0 & 1 & -4
\end{pmatrix},
\]
the process stops here, and where we expect to get an equation \(sb+tc=\gcd{b,c}\), instead we get
\[
0(-8)+1(-4)=-4, 
\]
which has the wrong sign, a negative greatest common divisor.
So if the answer pops out a negative for the greatest common divisor, we change signs: \(t=0, s=-1, \gcd{b,c}=4\).
\end{example}

\begin{theorem}\label{theorem:B\'ezout.euclidean}
The extended Euclidean algorithm calculates B\'ezout coefficients.
In particular, B\'ezout coefficients \(s,t\) exist for any integers \(b,c\) with \(c \ne 0\).
\end{theorem}
\begin{proof}
At each step in the extended Euclidean algorithm, the third column proceeds exactly by the Euclidean algorithm, replacing the larger number (in absolute value) with the remainder by the division, except for perhaps a minus sign.
So at the last step, the last nonzero number in the third column is the greatest common divisor, except for perhaps a minus sign.

Our extended Euclidean algorithm starts with
\[
\begin{pmatrix}
1 & 0 & b \\
0 & 1 & c
\end{pmatrix},
\]
a matrix which satisfies
\[
\begin{pmatrix}
1 & 0 & b \\
0 & 1 & c
\end{pmatrix}
\begin{pmatrix}
b \\
c \\
-1
\end{pmatrix}
=
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\]
It is easy to check that if some \(2 \times 3\) matrix \(M\) satisfies
\[
M
\begin{pmatrix}
b \\
c \\
-1
\end{pmatrix}
=
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\]
then so does the matrix you get by adding any multiple of one row of \(M\) to the other row.

Eventually we get a zero in the final column, say
\[
\begin{pmatrix}
s & t & d \\
S & T & 0
\end{pmatrix},
\]
or
\[
\begin{pmatrix}
s & t & 0 \\
S & T & D
\end{pmatrix}.
\]
It doesn't matter which since, at any step, we could always swap the two rows without changing the steps otherwise.
So suppose we end up at
\[
\begin{pmatrix}
s & t & d \\
S & T & 0
\end{pmatrix}.
\]
But
\[
\begin{pmatrix}
s & t & d \\
S & T & 0
\end{pmatrix}
\begin{pmatrix}
b \\
c \\
-1
\end{pmatrix}
=
\begin{pmatrix}
sb+tc-d \\
0
\end{pmatrix}.
\]
This has to be zero, so: \(sb+tc=d\).
\end{proof}


\begin{proposition}
Take two integers \(b,c\), not both zero.
The number \(\gcd{b,c}\) is precisely the smallest positive integer which can be written as \(sb+tc\) for some integers \(s,t\).
\end{proposition}
\begin{proof}
Let \(d \defeq \gcd{b,c}\).
Since \(d\) divides into \(b\), we divide it, say as \(b=Bd\), for some integer \(B\).
Similarly, \(b=Cd\) for some integer \(C\).
Imagine that we write down some positive integer as \(sb+tc\) for some integers \(s,t\).
Then \(sb+sc=sBd+tCd=(sB+tC)d\) is a multiple of \(d\), so no smaller than \(d\).
\end{proof}

\begin{sagesilent}
def bezpretty(b,c):
    p,q,r=1,0,b
    s,t,u=0,1,c
    M = matrix([[p,q,r],[s,t,u]])
    result="\\begin{align*}\n"
    result=result+"&{}\\\\ ".format(latex(M))
    while u<>0:
        if abs(r)>abs(u):
            p,q,r,s,t,u=s,t,u,p,q,r
        Q=u//r
        s,t,u=s-Q*p,t-Q*q,u-Q*r
        M = matrix([[p,q,r],[s,t,u]])
        result=result+"&{}\\\\ \n ".format(latex(M))
    result=result+"&({})({})+({})({})={}\\\\ \n ".format(p,b,q,c,r)
    result=result+"\\end{align*}"
    return result
\end{sagesilent}

\begin{problem}{euclidean.algorithm:try.Bezout}
Find B\'ezout coefficients and greatest common divisors of
\begin{enumerate}
\item 2468, 180
\item 79, -22
\item 45,16
\item -1000,2002
\end{enumerate}
\end{problem}
\begin{answer}{euclidean.algorithm:try.Bezout}
\begin{enumerate}
\item
\sagestr{bezpretty(2468,180)}
\item
\sagestr{bezpretty(79,-22)}
\item
\sagestr{bezpretty(45,16)}
\item
\sagestr{bezpretty(-1000,2002)}
\end{enumerate}
\end{answer}


\section{The greatest common divisor of several numbers}

Take several integers, say \(m_1, m_2, \dots, m_n\), not all zero.
To find their greatest common divisor, we only have to find the greatest common divisor of the greatest common divisors.
For example,
\[
\gcd{m_1, m_2, m_3}
=\gcd{\gcd{m_1, m_2}, m_3}
\]
and
\[
\gcd{m_1, m_2, m_3,m_4}
=\gcd{\gcd{m_1, m_2}, \gcd{m_3,m_4}}
\]
and so on.

\begin{problem}{euclidean.algorithm:multi.gcd}
Use this approach to find \(\gcd{54,90,75}\).
\end{problem}


\section{B\'ezout coefficients of several numbers}

Take integers \(m_1, m_2, \dots, m_n\), not all zero.
We will see that their greatest common divisor is the smallest positive integer \(d\) so that
\[
d = s_1 m_1 + s_2 m_2 + \dots + s_n m_n
\]
for some integers \(s_1, s_2, \dots, s_n\), the \emph{B\'ezout coefficients} of \(m_1, m_2, \dots, m_n\).
To find the B\'ezout coefficients, and the greatest common divisor:
\begin{enumerate}
\item
Write down the matrix:
\[
\begin{pmatrix}
1 & 0 & 0 & \dots & 0 & m_1 \\
0 & 1 & 0 & \dots & 0 & m_2 \\
0 & 0 & 1 & \dots & 0 & m_3 \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
0 & 0 & 0 & \dots & 1 & m_n
\end{pmatrix}
\]
\item
Subtract whatever integer multiples of rows from other rows, repeatedly, until the final column has exactly one nonzero entry.
\item
That nonzero entry is the greatest common divisor 
\[
d=\gcd{m_1, m_2, \dots, m_n}.
\]
\item
The entries in the same row, in earlier columns, are the B\'ezout coefficients, so the row looks like
\[
\begin{pmatrix}
s_1 & s_2 & \dots & s_n & d
\end{pmatrix}
\]
\end{enumerate}

\begin{example}
To find B\'ezout coefficients of \(12,20,18\):
\[
\begin{pmatrix}
1 & 0 & 0 & 12 \\
0 & 1 & 0 & 20 \\
0 & 0 & 1 & 18
\end{pmatrix}
\]
the smallest entry (in absolute value) in the final column is the 12, so we divide it into the other two entries:
\begin{enumerate}
\item
add \(-\text{row 1}\) to row 2, and 
\item
\(-\text{row 1}\) to row 3:
\end{enumerate}
\[
\begin{pmatrix}
1 & 0 & 0 & 12 \\
-1 & 1 & 0 & 8 \\
-1 & 0 & 1 & 6
\end{pmatrix}
\]
Now the smallest (in absolute value) is the \(6\), so we add integer multiples of its row to the two earlier rows:
\begin{enumerate}
\item
add \(-2\cdot\text{row 3}\) to row 1, and 
\item
\(-\text{row 3}\) to row 2:
\end{enumerate}
\[
\begin{pmatrix}
3 & 0 & -2 & 0 \\
0 & 1 & -1 & 2 \\
-1 & 0 & 1 & 6
\end{pmatrix}
\]
Now the \(2\) is the smallest \emph{nonzero} entry (in absolute value) in the final column.
Note: once a row in the final column has a zero in it, it becomes irrelevant, so we could save time by crossing it out and deleting it.
That happens here with the first row.
But we will leave it as it is, just to make the steps easier to follow.
Add \(-3\cdot\text{row 2}\) to row 3:
\[
\begin{pmatrix}
3 & 0 & -2 & 0 \\
0 & 1 & -1 & 2 \\
-1 & -3 & 4 & 0
\end{pmatrix}
\]
Just to carefully check that we haven't made any mistake, you can compute out that this matrix satisfies:
\[
\begin{pmatrix}
3 & 0 & -2 & 0 \\
0 & 1 & -1 & 2 \\
-1 & -3 & 4 & 0
\end{pmatrix}
\begin{pmatrix}
12 \\
20 \\
18 \\
-1
\end{pmatrix}
=
\begin{pmatrix}
0 \\
0 \\
0
\end{pmatrix}.
\]
In particular, our B\'ezout coefficients are the \(0,1,-1\), in front of the greatest common divisor, \(2\).
Summing up:
\[
0 \cdot 12 + 1 \cdot 20 + (-1)\cdot 18 = 2.
\]
\end{example}

\begin{problem}{euclidean.algorithm:multi.bezout}
Use this approach to find B\'ezout coefficients of \(54,90,75\).
\end{problem}

\begin{problem}{euclidean.algorithm:justify}
Prove that this method works.
\end{problem}

\begin{problem}{euclidean.algorithm:multi.lcm}
Explain how to find the least common multiple of three integers \(a,b,c\).
\end{problem}

\section{Sage}\label{section:euclid.sage}

\epigraph[author={Edsger Dijkstra}]{Computer science is no more about computers than astronomy is about telescopes.}\SubIndex{Dijkstra, Edsger}

The extended Euclidean algorithm is built in to sage:
\verb!xgcd(12,8)! returns \(\sage{xgcd(12,8)}\), the greatest common divisor and the B\'ezout coefficients, so that \(4=1 \cdot 12 + (-1) \cdot 8\).
We can write our own function to compute B\'ezout coefficients:
\begin{sageblock}
def bez(b,c):
    p,q,r=1,0,b
    s,t,u=0,1,c
    while u<>0:
        if abs(r)>abs(u):
            p,q,r,s,t,u=s,t,u,p,q,r
        Q=u//r
        s,t,u=s-Q*p,t-Q*q,u-Q*r
    return r,p,q
\end{sageblock}
so that \verb!bez(45,210)! returns a triple \((g,s,t)\) where \(g\) is the greatest commmon divisor of \(45,210\), while \(s,t\) are the B\'ezout coefficients.

Let's find B\'ezout coefficients of several numbers.
First we take a matrix \(A\) of size \(n \times (n+1)\).
Just as sage has lists \(L\) indexed starting at \verb!L[0]!, so sage indexes matrices as \(A_{ij}\) for \(i\) and \(j\) starting at zero, i.e. the upper left corner entry of any matrix \verb!A! is \verb!A[0,0]! in sage.
We search for a row which has a nonzero entry in the final column:
\begin{sageblock}
def row_with_nonzero_entry_in_final_column(A,rows,columns,starting_from_row=0):
    for i in range(starting_from_row,rows):
        if A[i,columns-1]<>0:
            return i
    return rows
\end{sageblock}
Note that we write \verb!starting_from_row=0! to give a \emph{default value}: we can specify a row to start looking from, but if you call the function \verb!rwo_with_nonzero_entry_in_final_column! without specifying a value for the variable \verb!starting_from_row!, then that variable will take the value \verb!0!.
We use this to find the B\'ezout coefficients of a list of integers:
\begin{sageblock}   
def bez(L):
    n=len(L)
    A=matrix(n,n+1)
    for i in range(0,n):
        for j in range(0,n):
            if i==j:
                A[i,j]=1
            else:
                A[i,j]=0
                A[i,n]=L[i]
    k=row_with_nonzero_entry_in_final_column(A,n,n+1)
    l=row_with_nonzero_entry_in_final_column(A,n,n+1,k+1)
    while l<n:
        if abs(A[l,n])>abs(A[k,n]):
            k,l=l,k
        q=A[k,n]//A[l,n]
        for j in range(0,n+1):
            A[k,j]=A[k,j]-q*A[l,j]
        k=row_with_nonzero_entry_in_final_column(A,n,n+1)
        l=row_with_nonzero_entry_in_final_column(A,n,n+1,k+1)
    return ([A[k,j] for j in range(0,n)],A[k,n])
\end{sageblock}
Given a list \(L\) of integers, we let \(n\) be the length of the list \(L\), form a matrix \(A\) of size \(n \times (n+1)\), put the identity matrix into its first \(n\) columns, and put the entries of \(L\) into its final column.
Then we find the first row \verb!k! with a nonzero entry in the last column.
Then we find the next row \verb!l! with a nonzero entry in the last column.
Swap \verb!k! and \verb!l! if needed to get row \verb!l! to have smaller entry in the last column.
Find the quotient \(q\) of those entries, dropping remainder.
Subtract off \(q\) times row \verb!l! from row \verb!k!.
Repeat until there is no such row \verb!l! left.
Return a list of the numbers.
Test it out:
\begin{sagesilent}
R=bez([-2*3*5,2*3*7,2*3*11])
\end{sagesilent}
\verb!bez([-2*3*5,2*3*7,2*3*11])! yields \(\sage{R}\), giving the B\'ezout coefficients as the list \(\sage{R[0]}\) and the greatest common divisor next: \(\sage{R[1]}\).


